### 2020-05-14 manuscript.Rmd
minor revisions an corrections in manuscript format tbl function kable extra table

parkrun, an international movement which organises free weekly 5km running events, might provide valuable lessons for public health professional with respect to both. Since its founding in 2004, as a small event in London with 13 participants, it has grown to become one of the world's largest mass sporting events, with up to 360,000 participants in more than 20 countries.[@ref] The volunteer-led running events are often characterised as accessible and inclusive.[@stevinson2015] The organisation has been widely praised as being successful in encouraging participation particularly in individuals who were previously inactive.[@stevinson2004; @wiltshire2018]

@@

  This study is a cross-sectional ecological analysis of the socio-economic disparities in geographic access to parkrun events in England at the end of 2018. All analyses were conducted on the level of Lower Layer Super Output Areas (LSOAs), which divide England into 32,844 geographic units with, on average, have a population of approximately 1,700. We assessed the relationship between access, defined as the distance (as the crow flies) to the nearest parkrun event, and socio-economic deprivation, measured using the Index of Multiple Deprivation. In addition, we used information on green spaces in England to conducted a simple location-allocation analysis, in order to identify 200 locations for future parkrun events, that maximise access for access for the population.

@@

  For this study, we combined data on three types of geospatial entities: 1) LSOAs, 2) parkrun event locations, and 3) public green spaces.

  1) The geographic locations of all 32,844 LSOAs, defined by the coordinates of its population-weighted centroid, 2017 total population estimates, and the 2015 Index of Multiple Deprivation (IMD) were retrieved from the Office for National Statistics [@ons2018; @ref; @ref].

  2) We included all 465 public parkrun events which were in operation in England by December 12th 2018 - on this date, Sport England announced that their plan to provide funding for setting up 200 additional parkrun events across England.[@ref] The coordinates of all 465 events were manually scrapped from the parkrun UK website.[@ref]

  3) The location coordinates of 143,822 public green spaces were retrieved from an open dataset of Ordnance Survey.[9] parkrun events are held in various settings and terrains, and not always require a 5km loop - some events have courses that involve running back and forth. After evaluating existing parkrun event courses, we decided to consider all public parks, gardens, and playing fields in England with an area of 0.1 km$^2$ (e.g. 316m x 316m) or more potentially suitable for hosting events (n=2,842).

@@

  Access to parkrun was measured in terms of the distance (as the crow flies) to the nearest event. For each of the 32,844 LSOAs, we computed the geodesic distances between its population-weighted centroid and all 465 parkun events and then selected the shortest distance (in km).

  The socio-economic deprivation of LSOAs was measured using IMD. It is a measure of measure of relative deprivation, which combines 37 indicators from seven domains (income, employment, education and skills, health and disability, crime, housing and services, and living environment) into a single score. The score ranges from 0 (least deprived) to 100 (most deprived).[@smith2015]

  Further covariates (e.g. population density or demographics), which are likely to affect the availability of parkrun events, were not taken into account, because we did not aim to assess to what extent deprivation independently 'explains' access. Rather, we sought to evaluate whether or not people living in deprived areas actually have better or worse access, given all other circumstances.

@@

  Mean, standard deviation, median, interquartile range, and range were used as descriptive statistics. The correlation between IMD and the distance to the nearest parkrun event on the LSOA level were investigated using the Pearson and Spearman method. The correlation coefficients were computed using the LSOAs' total population as weights. A scatter plot with a fitted linear regression model are presented to illustrate the relationship. Our hypothesis was that more deprived areas had worse access, i.e. longer distances to the nearest parkrun event than more affluent areas.

@@

  We assessed the overall impact of setting up 200 new parkrun events at the identified optimal location on the geographic access to parkrun events in England. We also conducted a distributional effectiveness analysis, in which we grouped LSOAs into IMD quintiles (most, more, median, less, least deprived) and subsequently assessed how setting up the 200 new events would affect access in each of those subgroups.

@@

  As of 12th December 2018, approximately 7%, 69%, and 91% of the English population lived within 1, 5, and 10km of a parkrun event. Only 578,043 people (1% of the English population) lived more than 20km from an event. The average (SD) and median (IQR) distance to the nearest parkrun event were 4.65 (4.22) and 3.39 (1.99-5.83). The worst access was observed for 2,259 people living on the Isles of Scilly, who live about 76.44km (as the crow flies) away from the next parkrun event on the mainland. Further descriptive statistics are provided in table 1.

| **Table 1: Descriptive statistics of LSOAs and parkrun events** |
| ---|---|---|--- |
| <td colspan=4>**LSOA (n =n 32,844) ** |
| ---|---|---|--- |
| |Mean (SD) | Median (Q1-Q3) | Range |
| Population | 1,693.44 (405.35) | 1,612 (1,452-1,834) | 362-13,404 |
| IMD score | 21.67 (15.59) | 17.40 (9.65-30.07) | 0.48-92.60 |
| Distance to the nearest parkrun event location | 4.65 (4.22) | 3.39 (1.99-5.83) | 0.04-76.44 |
|
| <td colspan=4>**Parkrun Events (n = 465)** |
| ---|---|---|--- |
| Served LSOAs\* | 70.63 (42.81) | 62 (40-87) | 6-350 |
| Served population\* | 119,612 (74,290) | 103,952 (68,837-151,488) | 7,855-628,010 |
| ---|---|---|--- |
 | \* Number of LSOAs or population for which a given parkrun event is the nearest. |

@@

The distributional analysis of distances by IMD quantile showed that people living in the 20% most deprived LSOAs  had the best geographic access, with an average and median distance to the nearest parkrun event of 3.51 and 2.79 km. Depending on the metric, the worst access was observed for LSOAs in the middle (mean distance = 3.36km) or the less deprived group (median distance = 3.93km). Further results of the distributional analysis are shown in Table 2 in the following section.

@@

Table 2 shows the difference in geographic access before and after the introduction of the 200 new parkrun events at the identified optimal green spaces. Results are presented for the overall effect and stratified by IMD quantiles. The 200 new events amplified the relationship between IMD and geographic access further: the population-weighted Pearson and Spearman correlation coefficients changed from -0.15 and -0.18 before, to -0.20 and -0.23 afterwards. Nevertheless, the improvement in access was smallest for LSOAs in the most deprived quantile.

@@

```{r}
# Table 2
# quantile(lsoa_sp$imd_sc,seq(0,1,0.2))
# cbind(mndst_tbl,mndst_new_tbl)
```

>>>>

parkrun, an international movement which organises free weekly 5km running events, might provide valuable lessons for public health professional with respect to both. Since its founding as a small event in London with 13 participants in 2004, it has grown to become one of the world's largest mass sporting events, with up to 360,000 participants in more than 20 countries.[@ref] The volunteer-led running events are often characterised as accessible and inclusive.[@stevinson2015] The organisation has been widely praised as being successful in encouraging participation particularly in individuals who were previously inactive.[@stevinson2004; @wiltshire2018]

@@

  This study is a cross-sectional ecological analysis of the socio-economic disparities in geographic access to parkrun events in England at the end of 2018. All analyses were conducted on the level of Lower Layer Super Output Areas (LSOAs), which divide England into 32,844 geographic units which, on average, have a population of approximately 1,700. We assessed the relationship between access, defined as the distance (as the crow flies) to the nearest parkrun event, and socio-economic deprivation, measured using the Index of Multiple Deprivation. In addition, we used information on public green spaces in England to conducted a simple location-allocation analysis, in order to identify 200 locations for future parkrun events, that maximise access for the population.

@@

  For this study, we combined data on three types of geospatial entities: 1) LSOAs, 2) parkrun events, and 3) public green spaces.

  1) For all 32,844 LSOAs, we retrieved geographic locations, defined by the coordinates of its population-weighted centroid, 2017 total population estimates, and the 2015 Index of Multiple Deprivation (IMD) from the Office for National Statistics [@ons2018; @ref; @ref].

  2) We included all 465 public parkrun events which were in operation in England by December 12th, 2018 - on this date, Sport England announced their plan to provide funding for setting up 200 additional parkrun events across England.[@ref] The locations of the events were manually scrapped from the parkrun UK website.[@ref]

  3) The locations of public green spaces in England were retrieved from an open dataset of Ordnance Survey.[9] Even though parkrun events do not always require a 5km loop (some courses involve running back and forth) and are held in various settings and terrains, not all green spaces are suitable to host running events. After evaluating existing parkrun event courses, we decided to consider all public parks, gardens, and playing fields with an area of 0.1 km$^2$ (e.g. 316m x 316m) or more potentially suitable for hosting events (n=2,842).

@@

  The two variables of interest were access to parkrun events and deprivation of LSOAs.

  Access to parkrun was measured in terms of the distance as the crow flies (in km) to the nearest event. For each of the 32,844 LSOAs, we computed the geodesic distances between its population-weighted centroid and all 465 parkun events that were in operation on December 12th, 2018, and then selected the shortest distance.

  The socio-economic deprivation of LSOAs was measured using the 2015 IMD. It is a measure of relative deprivation, which has been used in many similar studies. The IMD combines 37 indicators from seven domains (income, employment, education and skills, health and disability, crime, housing and services, and living environment) into a single score. The score ranges from 0 (least deprived) to 100 (most deprived).[@smith2015]

  Other covariates, which are likely to affect the availability of parkrun events within an area (e.g. population density or demographics), were not taken into account, because we did not aim to assess to what extent deprivation independently 'explained' access. Rather, we sought to evaluate whether or not people living in deprived areas have better or worse access, given the actual circumstances.

@@

  Mean, standard deviation, median, interquartile range, and range were used as descriptive statistics. We then assessed the association between the IMD and the distance to the nearest parkrun event on the LSOA level. Our hypothesis was that more deprived areas had worse access, i.e. longer distances to the nearest parkrun event than more affluent areas. Pearson and Spearman correlation coefficients were computed using the LSOAs' total population as weights. We also conducted a stratified analysis, for which we grouped LSOAs into IMD quintiles (most, more, median, less, least deprived) and assessed access to parkrun events in each strata.

@@

  We assessed the overall impact of setting up 200 new parkrun events at the identified optimal location on the geographic access to parkrun events in England. 
  In order to conduct a distributional analysis, we investigated the effect of setting up the 200 new events on LSOAs across IMD quintiles.

@@

  As of 12th December 2018, approximately 7%, 69%, and 91% of the English population lived within 1, 5, and 10km distance of a parkrun event. Only 578,043 people (1% of the English population) lived more than 20km from an event. The average (SD) and median (IQR) distance to the nearest parkrun event were 4.65 (4.22) and 3.39 (1.99-5.83). The worst access was observed for 2,259 people living on the Isles of Scilly, who live about 76.44km (as the crow flies) away from the next parkrun event on the mainland. Further descriptive statistics are provided in table 1.

```{r echo=FALSE, message=FALSE, warning=FALSE}
load(file = "./output/desc_tbls.RDS")
library(knitr)
library(kableExtra)

kable(rbind(desc_tbl1,desc_tbl2),
      caption = "Table 1: Descriptive statistics of LSOAs and parkrun events",
      format = "latex",
      longtable = T, 
      booktabs = T) %>%
  group_rows(group_label = "LSOAs (n = 32,844)",
             start_row = 1,hline_before=F,
             end_row = 3) %>%
   group_rows(group_label = "Parkrun events (n = 465)",
             start_row = 4,hline_before=T,
             end_row = 5) %>%
  add_footnote("foot1")

```

@@

The distributional analysis of distances by IMD quintile showed that people living in the 20% most deprived LSOAs  had the best geographic access, with an average and median distance to the nearest parkrun event of 3.51 and 2.79 km. Depending on the metric, the worst access was observed for LSOAs in the middle (mean distance = 3.36km) or the less deprived group (median distance = 3.93km). Further results of the distributional analysis are shown in Table 2 in the following section.

@@

Table 2 shows the difference in geographic access before and after the introduction of the 200 new parkrun events at the identified optimal green spaces. Results are presented for the overall effect and stratified by IMD quintiles. The 200 new events amplified the relationship between IMD and geographic access further: the population-weighted Pearson and Spearman correlation coefficients changed from -0.15 and -0.18 before, to -0.20 and -0.23 afterwards. Nevertheless, the improvement in access was smallest for LSOAs in the most deprived quintile.

@@

```{r}
# Table 2
# quintile(lsoa_sp$imd_sc,seq(0,1,0.2))
# cbind(mndst_tbl,mndst_new_tbl)
```
